Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-03-23 index.qmd
3/23 updates

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)

@@

where $h(u)$ is the hazard rate function [@klein2005]. This function
provides a better understanding of the survival function for graphing
purposes.

@@

To generate the CPH model with properly formatted survival data, a
survival object must first be created using the R package `survival`.
The survival object is created by running `Surv(time, event)` where
`time` is the time field and `event` is the event field.

During the analysis, multiple models will be created with a different
combination of covariates.

@@

This assumption can be tested using the R package `survival`. The
package contains a proportional hazards test called `cox.zph()`
[@R-base]. Using the hypothesis,

@@

the `cox.zph()` calculates the scaled Schoenfeld partial residuals and
returns the $\chi^2$ value, the degrees of freedom, and p-values for
each individual covariate along with a global value. Using an
$\alpha = 0.05$ and the global p-value, the determination of whether to
reject or fail to reject $H_0$ can be made. In addition to the
hypothesis test, these Shoenfeld residuals can be plotted against time
for each covariate to visualize the assumption. When plotting the
residuals against time, if the assumption is not broken it would be
expected that the points show roughly a straight horizontal line to
indicate constant hazard over time [@nahhas2025].

@@

Residuals can first be found by using the `resid(type = "martingale")`
function and passing martingale in the type argument creating a vector
of residuals that can be plotted against the range of values for each
covariate.

@@

This assumption describes how if an event is censored it should not be
considered when the model is being created. Censored data is difficult
to accommodate when Cox proportional hazards models are used in other
industries, such as the medical field, because observation times for the
subjects involved in studies may have decades before an event is
observed and the event may not have the opportunity to occur during the
study period. In this scenario, it would not be appropriate to consider
this subject when modeling survival.

@@

Survival times of observed subjects should be independent of each other.
In an example where survival of plane engines were being evaluated, if
each plane has two engines and the failure of one engine caused the
accelerated failure of another due to increased workload, it would
produce biased conclusions and violate the independence assumption.

@@

There are multiple different methods to visualize data when creating a
Cox proportional hazards model. Survival curves are a popular way to
show survival probabilities over time, but there are other
visualizations that are helpful when interpreting the results of the
model.

@@

A forest plot will be created using the `plot_model()` function from the
sjPlot package. The CPH model can be passed as the sole argument to
visualize the hazard ratios. An example forest plot is shown below
plotting the hazard ratio for each covariate in the model. The point
represents the calculated hazard ratio and the whiskers show the 95%
confidence interval.

@@

After generating the CPH model, a Kaplan-Meier survival curve can be
generated to view survival probabilities against time. These plots are
important when modeling survival data because they clearly demonstrate
the time where the event being modeled for is expected to occur. ' The
survival curve will be plotted by using the survival package's
`survfit()` function. The object created will be plotted using `plot()`

Below is an example survival curve showing survival probabilities over
time. The line represents the median survival probability with the red
area around it showing the 95% confidence interval for the values. The
curve can be interpreted as showing the probability of event would be
expected to happen around a time of 300 where the survival probability
is below 50%.

@@

Cox proportional hazards models can produce an assortment of useful
metrics to interpret the model. The model created can be used to
estimate risk of the event occurring, rate of occurrence, survival
probability, and time to event.

Survival probabilities can be calculated in R by using
`stats::predict()` after creating a CPH model. This function will be
used to create a column showing the survival probability for each row in
the testing data.

@@

@@

Below is a univariate test to determine which covariates to use in the
model

@@

```{r}
cox <- coxph(Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5,
             data = train,
             id = id)

summary(cox)
```

#### Checking Assumptions

**Checking Proportional Hazards Assumption**

```{r}
cox.zph(cox)

plot(cox.zph(cox))

@@

**Checking Linearity Assumption**

Example for one predictor

```{r}
martingale <- resid(cox, type = "martingale")

@@

**Checking Independent Censoring**

@@

```{r}
plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

@@

```{r}
survfit <- survfit(cox)

@@

```{r}
predictiondata <- test %>%
  group_by(id) %>%
  slice_max(time) %>%
  select(id, time, os1, os3, sm1, sm2, sm5, status)

coef <- data.frame(Variable = c("Z1", "Z2", "Z3", "Z4", "Z5"),
           CoefficientValue = cox$coefficients)
kable(coef)
```

@@

```{r}
test$hazardrates <- predict(cox,
        newdata = test,
        type = "risk")

@@

```{r}
test$survprob <- predict(cox, type = "survival", newdata = test)

@@

>>>>

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)

@@

where $h(u)$ is the hazard rate function [@klein2005]. This function
provides a better understanding of the survival function for graphing
purposes.[@Amini2015]

@@

During the analysis, multiple models will be created with a different
combination of covariates. Covariates will be considered in the models that pass the covariate test. After creating the models, the assumptions described below will be tested. Out of the multiple models created, models will also not be considered where the assumptions are not true.

The model used through the rest of the analysis to make predictions will be the model with the best evaluation where the assumptions ar true.

@@

This assumption can be tested with a hypothesis test using the hypothesis:

@@

The `cox.zph()` function calculates the scaled Schoenfeld partial residuals and
returns the $\chi^2$ value, the degrees of freedom, and p-values for
each individual covariate along with a global value. Using an
$\alpha = 0.05$ and the global p-value, the determination of whether to
reject or fail to reject $H_0$ can be made. In addition to the
hypothesis test, these Shoenfeld residuals can be plotted against time
for each covariate to visualize the assumption. When plotting the
residuals against time, if the assumption is not broken it would be
expected that the points show roughly a straight horizontal line to
indicate constant hazard over time [@nahhas2025].

@@

@@

This assumption describes how if an event is censored it should not be
considered when the model is being created. Censored data is difficult
to accommodate when Cox proportional hazards models are used in other
industries, such as the medical field, because observation times for the
subjects involved in studies may have decades before an event is
observed and the event may not have the opportunity to occur during the
study period. In this scenario, it would not be appropriate to consider
this subject when modeling survival.[@Amini2015]

@@

Survival times of observed subjects should be independent of each other.
In an example where survival of plane engines were being evaluated, if
each plane has two engines and the failure of one engine caused the
accelerated failure of another due to increased workload, it would
produce biased conclusions and violate the independence assumption.[@Amini2015]

@@

### Model Visualization and Prediction

There are multiple different methods to visualize data when creating a
Cox proportional hazards model. Survival curves are a popular way to
visualize survival probabilities over time, but there are other
visualizations that are helpful when interpreting the results of the
model.

Cox proportional hazards models can produce an assortment of useful
metrics to interpret the model as described by the equations above. The model created can be used to
estimate risk of the event occurring, rate of occurrence, survival
probability, and time to event.

@@

An example forest plot is shown below
plotting the hazard ratio for each covariate in the model. The point
represents the calculated hazard ratio and the whiskers show the 95%
confidence interval.

@@

After generating the CPH model, a Kaplan-Meier survival curve can be
generated to view survival probabilities against time. These plots are
important when modeling survival data because they clearly demonstrate
the time where the event being modeled for is expected to occur.

Below is an example survival curve showing survival probabilities over
time. The line represents the median survival probability with the red
area around it showing the 95% confidence interval for the values. The
curve can be interpreted as showing the probability of event would be
expected to happen around a time of 300 where the survival probability
is below 50%.[@Kuitunen2021]

@@

@@

Below are tables showing the structure of the training and testing data after the transformations describd above.

@@

Below is a univariate test to determine which covariates to use in the
model.

@@

To generate the CPH model, a
survival object must first be created using the R package `survival`.
The survival object is created by running `Surv(time, event)` where
`time` is the time field and `event` is the event field.

```{r}
cox1 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox2 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5)

cox3 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)

cox4 <- coxph(data = train, 
             Surv(time, status) ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox5 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )

cox6 <- coxph(data = train, 
             Surv(time, status) ~ os1  + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)

cox7 <- coxph(data = train, 
             Surv(time, status) ~ os1  + sm13 + sm14 + sm19 + sm20)

cox8 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm21)

cox9 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm2 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)

cox10 <- coxph(data = train, 
             Surv(time, status) ~  os3 + sm1 + sm2 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)

@@

The resulting models and their corresponding concordance index values are below.

```{r}
concordance_results <- data.frame(row.names = c(
  "Model 1",
  "Model 2",
  "Model 3",
  "Model 4",
  "Model 5",
  "Model 6",
  "Model 7",
  "Model 8",
  "Model 9",
  "Model 10"),
  Concordance = c(
  cox1$concordance[6],
  cox2$concordance[6],
  cox3$concordance[6],
  cox4$concordance[6],
  cox5$concordance[6],
  cox6$concordance[6],
  cox7$concordance[6],
  cox8$concordance[6],
  cox9$concordance[6],
  cox10$concordance[6]))

kable(concordance_results)

The analysis will proceed using model 1 because of the high concordance index. The full summary of model 1 is below.

```{r}
summary(cox1)

### Checking Assumptions

Despite the high concordance value for model 1, the assumptions will need to be evaluated before using the model to make predictions. If any of the assumptions for the model are not true, the model will need to be modified

#### Checking Proportional Hazards Assumption
The `cox.zph` function can be used to check the proportional hazards assumption. As mentioned above, the function calculates p values for each covariate. If the p values are < .05, the null hypothesis is rejected indicating the proportional hazards assumption is true. If the proportional hazards assumption is not true for any covariate, the model will be recreated without those covariates.

```{r}
cox.zph(cox1)
```
The output above shows the p values for the hypothesis test.

In addition to the hypothesis test, the Shoenfeld residuals calculated by the `cox.zph()` function are plotted against time below.

```{r}
# Function to put ph_plots on one visual
create_ph_plots <- function(coxmodel){
  # Function takes argument that is a cox model from coxph()

  phresults <- cox.zph(coxmodel)

  # Define dimensions of visual
  par(mfrow = c(3,5),
      mar = c(2,2,4,1))

  # Loop through each covariate creating a plot and adding it to the phplot list
  for (i in 1:(length(phresults$table[,1])-1)){
    phplots <- list()
    phplots[i] <- plot(phresults[i], # Create plot for each iteration
                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table

  }
  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
  par(mfrow = c(1,1)) # Reset plotting area
}

create_ph_plots(cox1)
```

The proportional hazards test shows several covariates do not meet the proportional hazards assumption.

The model is recreated below with another proportional hazards test.

```{r}

cox1.1 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
cox.zph(cox1.1)

The hypothesis test shows the covariates sm1, sm2, sm6, sm7, sm11, sm12, sm20, and sm21 do not meet the proportional hazards assumption. The model is created again below with a second hypothesis test showing all of the covariates pass the hypothesis test.

#### Checking Linearity Assumption

Residuals can first be found by using the `resid(type = "martingale")`
function and passing martingale in the type argument creating a vector
of residuals that can be plotted against the range of values for each
covariate.
Example for one predictor

```{r}
martingale <- resid(cox1.1, type = "martingale")

@@

#### Checking Independent Censoring

@@

A forest plot will be created using the `plot_model()` function from the
sjPlot package. The CPH model can be passed as the sole argument to
visualize the hazard ratios.

```{r}
plot_model(cox1.1,dot.size = 4,line.size = 2,colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

@@

 The
survival curve will be plotted by using the survival package's
`survfit()` function. The object created will be plotted using `plot()`
```{r}
survfit <- survfit(cox1.1)

@@

```{r}
#predictiondata <- test %>%
  #group_by(id) %>%
  #slice_max(time) %>%
  #select(id, time, os1, os3, sm1, sm2, sm5, status)

#coef <- data.frame(Variable = c("Z1", "Z2", "Z3", "Z4", "Z5"),
           #CoefficientValue = cox1.1$coefficients)
#kable(coef)
```

@@

```{r}
test$hazardrates <- predict(cox1.1,
        newdata = test,
        type = "risk")

@@

```{r}
test$survprob <- predict(cox1.1, type = "survival", newdata = test)

@@

```{r}
